---
title: "Secsse performance"
author: "Thijs Janzen"
date: "2023-07-10"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Secsse performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 7)
knitr::opts_chunk$set(fig.height = 5)
library(secsse)
library(ggplot2)
```

secsse has gone over many versions since it's first appearance on CRAN in 2019, with various rates of computational performance. 
Here, we would like to shortly go over the main versions of secsse, and compare
their computational performance.

## Secsse Versions

### 1.0.0
The first version of secsse appeared in January of 2019 on CRAN. It used the
package deSolve to solve all integrations, and could switch between either using
a fully R based evaluation, or use FORTRAN to speed up calculations.
Furthermore, using the foreach package, within-R parallelization was
implemented. However, parallelization only situationally improved computation
times, and generally, computation was relatively slow.

### 2.0.0
Version 2.0.0 appeared in June of 2019 on CRAN and extended the package with the
cla framework, e.g. including state shifts during speciation / asymmetric 
inheritance during speciation. 

### 2.5.0
Version 2.5.0 appeared in 2021 on GitHub and was published in May 2023 on CRAN.
Version 2.5.0 marks the first version using C++ to perform the integration,
and it used tbb (from the RcppParallel package) to perform multithreading. This
marks a ten fold increase in speed over previous versions.

### 2.6.0
Version 2.6.0 appeared on CRAN in July 2023, and introduced many functions 
suited to prepare the parameter structure for secsse. It also introduced a new
C++ code base for the standard likelihood, making smarter use of
parallelization, this marks another 10-fold increase in speed.

### 3.0.0
Version 3.0.0 is expected to arrive to CRAN in the second half of 2023. It 
extends the C++ code base used for the standard likelihood to the cla
likelihood, harnessing the same computation improvement. 

## Speed
Using a standardized computation test of calculating the likelihood of a system
with two observed and two concealed traits, on a tree of ~500 tips we calculated
the computation time using either the cla or the standard likelihood. Loading 
and reloading different versions of the same package inevitably requires
restarting R in between to clear cache memory and avoid using parts of code not 
completely unloaded. Hence, here we do not actually perform the benchmark, but 
load the results directly from file:

```{r plot_results}
load("timing_data.RData")

ggplot(timing_data, aes(x = version, y = time, col = as.factor(num_threads))) +
  geom_boxplot() +
  scale_y_log10() +
  xlab("secsse version") +
  ylab("Computation time (seconds)") +
  labs(col = "Number of\nthreads") +
  theme_classic() +
  scale_color_brewer(type = "qual", palette = 2) +
  facet_wrap(~type)
```

It is clear that we have come a long way since 2019, and that current versions
of secsse are approximately a factor 100 faster. Note that for the cla
likelihood, there are not timings available for version 1.0.0, because that 
version did not contain the cla likelihood versions yet. 

## Appendix
### Testing code standard likelihood
```{r standard likelihood}

run_this_code <- FALSE
if (run_this_code) {
  set.seed(42)
  out <- DDD::dd_sim(pars = c(0.5, 0.3, 1000), age = 30)
  phy <- out$tes
  cat("this tree has: ", phy$Nnode + 1, " tips\n")
  
  traits <- sample(c(0, 1), ape::Ntip(phy), replace = TRUE)
  b <- c(0.04, 0.04)  # lambda
  d <- rep(0.01, 2)
  userTransRate <- 0.2 # transition rate among trait states
  num_concealed_states <- 2
  sampling_fraction <- c(1, 1)
  toCheck <- secsse::id_paramPos(traits,num_concealed_states)
  toCheck[[1]][] <- b
  toCheck[[2]][] <- d
  toCheck[[3]][,] <- userTransRate
  diag(toCheck[[3]]) <- NA
  root_state_weight <- "proper_weights"
  use_fortran <- TRUE
  methode <- "odeint::bulirsch_stoer"
  cond <- "noCondit"
  
  # the different secsse versions have similar, but not identical 
  # syntax (mainly, they handle multi-threading / parallelization different)
  run_secsse_new <- function(nt) {
    secsse::secsse_loglik(parameter = toCheck,
                          phy = phy,
                          traits = traits,
                          num_concealed_states = num_concealed_states,
                          cond = cond,
                          root_state_weight = root_state_weight,
                          sampling_fraction = sampling_fraction,
                          num_threads = nt,
                          is_complete_tree = FALSE)
  }
  
  run_secsse_old <- function(use_parallel) {
    secsse::secsse_loglik(parameter = toCheck,
                          phy = phy,
                          traits = traits,
                          num_concealed_states = 
                            num_concealed_states,
                          sampling_fraction = sampling_fraction,
                          run_parallel = use_parallel)
  }
  
  measure_time <- function(local_fun, num_repl, parallel) {
    vv <- c()
    for (r in 1:num_repl) {
      t1 <- Sys.time()
      local_fun(parallel)
      t2 <- Sys.time()
      vv[r] <- difftime(t2, t1, units = "secs")
    }
    return(vv)
  }
  
  if (packageVersion("secsse") < 2.5) {
    t1 <- measure_time(run_secsse_old, 10, FALSE)
    t2 <- measure_time(run_secsse_old, 10, TRUE)
    to_add <- cbind(t1, as.character(packageVersion("secsse")), 1)
    to_add2 <- cbind(t2, as.character(packageVersion("secsse")), 2)
    timing_data <- rbind(timing_data, to_add, to_add2)
  } else {
    t1 <- measure_time(run_secsse_new, 10, 1)
    t2 <- measure_time(run_secsse_new, 10, 2)
    t3 <- measure_time(run_secsse_new, 10, 8)
    to_add <- cbind(t1, as.character(packageVersion("secsse")), 1)
    to_add2 <- cbind(t2, as.character(packageVersion("secsse")), 2)
    to_add3 <- cbind(t3, as.character(packageVersion("secsse")), 8)
    timing_data <- rbind(timing_data, to_add, to_add2, to_add3)
  }
}

```

### Testing code Cla likelihood
```{r testing_cla}
run_code <- FALSE
if (run_code) {
  set.seed(42)
  #set.seed(51)
  out <- DDD::dd_sim(pars = c(0.5 , 0.3, 1000), age = 30)
  phy <- out$tes
  cat("this tree has: ", phy$Nnode + 1, " tips and ", phy$Nnode, " internal nodes\n")
  
  num_concealed_states <- 3
  
  traits <- sample(c(0,1, 2), ape::Ntip(phy),replace = TRUE)
  
  sampling_fraction = c(1, 1, 1)
  idparlist <- cla_id_paramPos(traits, num_concealed_states)
  lambda_and_modeSpe <- idparlist$lambdas
  lambda_and_modeSpe[1,] <- c(0.2, 0.2, 0.2, 0.4, 0.4, 0.4, 0.01, 0.01, 0.01)
  
  parameter <- list()
  parameter[[1]] <- prepare_full_lambdas(traits, num_concealed_states,
                                         lambda_and_modeSpe)
  
  parameter[[2]] <- rep(0.05,9)
  
  masterBlock <- matrix(0.07, ncol = 3, nrow = 3, byrow = TRUE)
  diag(masterBlock) <- NA
  parameter[[3]] <- q_doubletrans(traits, masterBlock, diff.conceal = FALSE)
  
  run_secsse_new <- function(nt) {
    secsse::cla_secsse_loglik(parameter = parameter,
                              phy = phy,
                              traits = traits,
                              num_concealed_states = num_concealed_states,
                              sampling_fraction = sampling_fraction,
                              is_complete_tree = FALSE,
                              num_threads = nt,
                              atol = 1e-8,
                              rtol = 1e-6)
  }
  
  run_secsse_old <- function(use_parallel) {
    secsse::cla_secsse_loglik(parameter = parameter,
                              phy = phy,
                              traits = traits,
                              num_concealed_states = 
                                num_concealed_states,
                              sampling_fraction = sampling_fraction,
                              run_parallel = use_parallel)
  }
  
  measure_time <- function(local_fun, num_repl, parallel) {
    vv <- c()
    for (r in 1:num_repl) {
      t1 <- Sys.time()
      local_fun(parallel)
      t2 <- Sys.time()
      vv[r] <- difftime(t2, t1, units = "secs")
    }
    return(vv)
  }
  
  if (packageVersion("secsse") < 2.5) {
    t1 <- measure_time(run_secsse_old, 10, FALSE)
    t2 <- measure_time(run_secsse_old, 10, TRUE)
    to_add <- cbind(t1, as.character(packageVersion("secsse")), 1)
    to_add2 <- cbind(t2, as.character(packageVersion("secsse")), 2)
    timing_data <- rbind(timing_data, to_add, to_add2)
  } else {
    t1 <- measure_time(run_secsse_new, 10, 1)
    t2 <- measure_time(run_secsse_new, 10, 2)
    t3 <- measure_time(run_secsse_new, 10, 8)
    to_add <- cbind(t1, as.character(packageVersion("secsse")), 1)
    to_add2 <- cbind(t2, as.character(packageVersion("secsse")), 2)
    to_add3 <- cbind(t3, as.character(packageVersion("secsse")), 8)
    timing_data <- rbind(timing_data, to_add, to_add2, to_add3)
  }
}

```
