---
title: "Setting up a secsse analysis"
author: "Thijs Janzen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setting up a secsse analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setting up

When preparing a secsse analysis, it can be daunting to prepare the different required matrices and settings in order to be able to perform a meaningful analysis. Starting with secsse package version 2.6, there are now general helper functions available that can prepare all matrices for some general cases. Often, these general cases can already be applicable, alternatively, they can be modified later on to better reflect the intricacies of the specific studied system.

## Requirements for secsse analysis

To perform a secsse analysis, we want to use maximum likelihood to find the most likely values for our parameters, given a phylogenetic tree and tip states. To do so, secsse requires the user to specify how speciation changes the state of the daughter species in relation to the parent species, and requires the user to specify the number of unique speciation rates to be fitted. Here, we will explore a basic example.

### Two observed states, two hidden state
We start with a straightforward, simple case where we have two observed states (perhaps the presence / absence of an ornament or so), and we assume that the concealed state follows a similar structure, e.g. it also has two unique states. 
Now, we can specify three different models, 1) constant-rates model, where rates are not dependent on any trait, 2) Examined-Trait-Diversification (ETD), where rates are dependent on the observed trait and 3) CTD (Concealed-Trait-Diversification), where rates are dependent on the concealed trait.

#### Lambdas
To create the required lambda-matrices, we need as input information about the observed state names, the number of concealed states, and a transition_list object, which is a matrix defining the traits of daughter species upon speciation and their associated rate. We will here generate a default transition_list, but the user is free to create (and encouraged) one manually her/him self in order to reflect the focal system better.
We assume here that we have a trait with lables "S" and "N", and use the default settings:

```{r default_trans_list}
focal_list <- secsse::create_default_transition_list(state_names = c("S", "N"))
focal_list
```

With this list generated, we can now use this to populate our lambda matrices, using a constant rates model:
```{r default lambda matrices}
lambda_matrices <- secsse::create_lambda_matrices(state_names = c("S", "N"),
                                                  num_concealed_states = 2,
                                                  transition_list = focal_list,
                                                  model = "CR")
lambda_matrices
```

We see that there are four lambda matrices, one for each of the combined states (e.g. for each combination of observed and hidden states). So in this case we have our two observed states S and N, and the two hidden states A and B. This results in the four real states SA, NA, SB and NB. 

#### Extinction

We also need to specify an extinction rate:

```{r adding extinction}
mus <- secsse::create_mus(state_names = c("S", "N"),
                          num_concealed_states = 2,
                          model = "CR",
                          lambdas = lambda_matrices)
mus
```

#### Q matrix

To specify a q-matrix, we again need to specify the transitions using a transition list. Again, we use the standard settings.

```{r default_trans} 
q_list <- secsse::create_default_q_list(state_names = c("S", "N"),
                                        num_concealed_states = 2,
                                        mus = mus)

q_list

trans_matrix <- secsse::create_transition_matrix(state_names = c("S", "N"),
                                                 num_concealed_states = 2,
                                                 transition_list = q_list)
trans_matrix
```
Here, we find transitions from A->B, B->A but also S->N and N->S.

### Simulating data
Now, we can use our settings to perform an analysis.
Because we are lacking empirical data in this example, we will simulate a tree for this. 
To do so, we first need to specify our focal rates, and then fill them in.

```{r fill in parameters}

speciation <- 0.5
extinction <- 0.0
q_ab <- 0.5
q_ba <- 0.5
sp_ns <- 0.2
sp_sn <- 0.2

params <- c(speciation,
            extinction,
            sp_ns, sp_sn, q_ab, q_ba)

lambda_matrices_p <- secsse::fill_in(lambda_matrices,
                                     params)
trans_matrix_p <- secsse::fill_in(trans_matrix,
                                  params)
mus_p <- secsse::fill_in(mus,
                         params)
```

With the values replaced, we can now simulate an "empirical" dataset:

```{r simulate tree}
simulated_tree <- secsse::secsse_sim(lambdas = lambda_matrices_p,
                                     mus = mus_p,
                                     qs = trans_matrix_p,
                                     num_concealed_states = 2,
                                     crown_age = 5,
                                     conditioning = "obs_states",
                                     verbose = TRUE)
sim_traits <- simulated_tree$obs_traits
focal_tree <- simulated_tree$phy
```

### Maximum Likelihood

Given this data, we can now perform our maximum likelihood analysis. Here, we
choose to initialize our parameters with random values in [0, 1], we use 
multithreading to speed up the analysis, and use the subplex optimization 
method, as this has shown to be more reliable.

```{r maximum likelihood}
param_posit <- list()
param_posit[[1]] <- lambda_matrices
param_posit[[2]] <- mus
param_posit[[3]] <- trans_matrix

initpars <- params
initpars <- initpars[-2]



answ <- secsse::cla_secsse_ml(phy = focal_tree,
                              traits = sim_traits,
                              num_concealed_states = 2,
                              idparslist = param_posit,
                              idparsopt = c(1, 3, 4, 5, 6),
                              initparsopt = initpars,
                              idparsfix = c(0, 2),
                              parsfix = c(0.0, 0.0),
                              sampling_fraction = c(1, 1),
                              optimmethod = "subplex",
                              verbose = FALSE,
                              num_threads = 6)
```

We can now extract our parameters to get them in the right place:

```{r extract_pars}
found_pars_vals <- secsse::extract_par_vals(param_posit, answ$MLpars)
found_pars_vals
```

## Comparing models using AIC

We have done this now only for the CR model, but we can also use the 
CTD and ETD model. Let's do that semi-automagically! We first define a generic
function to optimize for a model:
```{r define_model_function}
fit_model <- function(tree, traits, model) {
  focal_list <- secsse::create_default_transition_list(state_names =
                                                         c("S", "N"))
  lambda_matrices <- secsse::create_lambda_matrices(state_names = c("S", "N"),
                                                    num_concealed_states = 2,
                                                    transition_list =
                                                        focal_list,
                                                    model = model)
  mus <- secsse::create_mus(state_names = c("S", "N"),
                            num_concealed_states = 2,
                            model = model,
                            lambdas = lambda_matrices)
  q_list <- secsse::create_default_q_list(state_names = c("S", "N"),
                                          num_concealed_states = 2,
                                          mus = mus)

  trans_matrix <- secsse::create_transition_matrix(state_names = c("S", "N"),
                                                   num_concealed_states = 2,
                                                   transition_list = q_list)

  param_posit <- list()
  param_posit[[1]] <- lambda_matrices
  param_posit[[2]] <- mus
  param_posit[[3]] <- trans_matrix

  max_indicator <- max(trans_matrix, na.rm = TRUE)

  # we cheat a bit by setting extinction to zero -
  # in a real analysis this should be avoided.
  extinct_rates <- unique(mus)
  idparsopt <- 1:max_indicator
  idparsopt <- idparsopt[-extinct_rates]
  idparsfix <- c(0, extinct_rates)
  parsfix <- rep(0.0, length(idparsfix))

  initpars <- runif(n = length(idparsopt))

  testthat::expect_output( # suppress output
  answ <- secsse::cla_secsse_ml(phy = focal_tree,
                                traits = traits,
                                num_concealed_states = 2,
                                idparslist = param_posit,
                                idparsopt = idparsopt,
                                initparsopt = initpars,
                                idparsfix = idparsfix,
                                parsfix = parsfix,
                                sampling_fraction = c(1, 1),
                                optimmethod = "subplex",
                                verbose = FALSE,
                                num_threads = 6)
  )
  found_pars_vals <- secsse::extract_par_vals(param_posit, answ$MLpars)
  aic <- 2 * max_indicator - 2 * as.numeric(answ$ML)
  return(list(pars = found_pars_vals,
              ml = as.numeric(answ$ML),
              aic = aic))
}
```

And then we loop over the different models:

```{r model looping}
found <- c()
for (focal_model in c("CR", "CTD", "ETD")) {
  local_answ <- fit_model(tree = focal_tree,
                          traits = sim_traits,
                          model = focal_model)
  found <- rbind(found, c(focal_model, local_answ$ml, local_answ$aic))
}
colnames(found) <- c("model", "LL", "AIC")
found <- as.data.frame(found)
found$LL <- as.numeric(found$LL)
found$AIC <- as.numeric(found$AIC)
found
```

Because we have simulated the tree using the CR model, we expect the model with
the lowest AIC to be the CR model again, and indeed we do find this!
